---
title: "Understanding Simpson's Paradox"
subtitle: "A Data Visualization Journey with Penguin Bill Dimensions"
format: gfm
output-file: README.md
execute:
  echo: true
  eval: true
---

## Introduction to Simpson's Paradox

**Simpson's Paradox** is one of the most fascinating and dangerous phenomena in data analysis. It occurs when a trend appears in different groups of data but disappears or reverses when these groups are combined. This paradox can lead to completely wrong conclusions if we don't look beyond surface-level statistics.

The classic example comes from UC Berkeley's 1973 graduate admissions data, where women appeared to be discriminated against in the overall admission rates, but when examined by department, women actually had equal or higher admission rates than men.

In this tutorial, we'll explore Simpson's Paradox using the beloved Palmer Station penguin dataset, examining the relationship between bill length and bill depth across different penguin species.

```{r setup}
#| message: false
#| warning: false
# Set CRAN mirror for package installation
options(repos = c(CRAN = "https://cran.rstudio.com/"))

# Install required packages if not already installed
if (!require(patchwork, quietly = TRUE)) {
  install.packages("patchwork")
  library(patchwork)
}

library(ggplot2)
library(palmerpenguins)
data(penguins)
```

## The Data Story: Penguin Bill Dimensions

Let's start by examining our data and building up to the paradox step by step.

### A. The Overall Relationship

```{r}
#| message: false
#| warning: false
plot1 <- ggplot(data = penguins,
        aes(x = bill_length_mm,
        y = bill_depth_mm)) +
  theme_minimal(16) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(title = "Overall Relationship: Bill Length vs. Bill Depth",
       subtitle = "Ignoring species differences",
       x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = 0, face= "italic"),
        plot.caption.position = "plot") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5)

plot1
```

At first glance, we see a **positive relationship** between bill length and bill depth. The red trendline suggests that longer bills tend to be deeper. This seems straightforward, but let's dig deeper...

### B. The Hidden Truth: Species-Specific Relationships

Now let's color our points by species and see what happens when we examine the relationship within each group:

```{r}
#| message: false
#| warning: false
#| eval: true
#| fig-cap: "Simpson's Paradox Revealed: Overall vs. Species-Specific Relationships"

plot2 <- ggplot(data = penguins,
       aes(x = bill_length_mm,
           y = bill_depth_mm,
           color = species)) +
  theme_minimal(16) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("darkorange","purple","cyan4"),
                     name = "Species") +
  labs(title = "Simpson's Paradox Revealed!",
       subtitle = "The relationship reverses when we consider species",
       x = "Bill length (mm)",
       y = "Bill depth (mm)",
       caption = "Red line: Overall trend | Colored lines: Within-species trends") +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = 0, face = "italic"),
        plot.caption.position = "plot",
        legend.position = "bottom") +
  # Overall trend (ignoring species)
  geom_smooth(method = "lm", se = FALSE, color = "red", 
              linewidth = 1.5, alpha = 0.8) +
  # Species-specific trends
  geom_smooth(method = "lm", se = FALSE, linewidth = 1)

plot2
```

**The Paradox Revealed!** 🎯

Notice what just happened:
- **Overall trend (red line):** Positive relationship - longer bills are deeper
- **Within each species (colored lines):** Negative relationship - longer bills are actually shallower!

This is Simpson's Paradox in action. The species variable is a **confounding variable** that completely reverses the apparent relationship when we aggregate the data.

### C. Side-by-Side Comparison

**🎯 Your Task:** Use Cursor's AI assistant to create a compelling side-by-side comparison plot!

**Instructions:**
1. Ask Cursor: *"Create a side-by-side comparison plot using patchwork that combines plot1 and plot2 with a title, subtitle, and caption explaining Simpson's Paradox"*
2. The plot should clearly show the contrast between the overall relationship and the species-specific relationships
3. Use `theme_minimal(16)` for consistency
4. **Replace these instructions with the actual plot code that Cursor generates**
5. **Render the README.md file** using `quarto render README.qmd`
6. Make sure to sync your changes to your GitHub folder for grading!

**What you should see:** A powerful visualization that makes Simpson's Paradox crystal clear - the same data telling two completely different stories depending on whether we consider the species grouping variable.

## Key Takeaways: Lessons from Simpson's Paradox

### The Power of Visualization

- **Always visualize your data** before drawing conclusions from statistical models
- A single trendline can be dangerously misleading when it masks underlying group differences
- **P-values are not the answer** - statistical significance doesn't guarantee the relationship is meaningful or correctly interpreted

### Simpson's Paradox Dangers

- **Aggregated data can reverse relationships** - what appears to be a positive correlation overall might be negative within each group
- **Business implications:** Making decisions based on aggregate trends without considering subgroups can lead to costly mistakes
- **The "lurking variable" problem:** Always ask "What am I missing?" when relationships seem counterintuitive

### Real-World Examples

- **UC Berkeley admissions (1973):** Overall admission rates suggested gender bias against women, but within each department, women had equal or higher admission rates
- **Baseball batting averages:** A player can have a higher batting average than another in both halves of a season, yet a lower overall average
- **Medical studies:** A treatment can appear harmful overall but beneficial within each age group
- **Marketing campaigns:** A campaign might seem ineffective overall but highly successful within specific customer segments

### Best Practices for Data Analysis

- **Stratify your analysis** by relevant grouping variables (demographics, categories, time periods)
- **Look for confounding variables** that might explain apparent relationships
- **Use multiple visualization approaches** (overall vs. faceted plots, side-by-side comparisons)
- **Question your assumptions** - if a relationship seems too good (or bad) to be true, it might be!

## Conclusion

Simpson's Paradox teaches us that **context matters**. The penguin data shows us that what appears to be a simple positive relationship between bill length and depth is actually masking a more complex reality where the relationship is negative within each species. This paradox reminds us to always dig deeper, consider confounding variables, and never trust aggregate statistics without examining the underlying groups.

Remember: **The truth is in the details, not in the summary statistics.**

